Exploratory Analysis of Superstore

### 2024-04-22 R/eda-code/eda_superstore.qmd
Project Part 5

---
title: "Exploratory Analysis of Superstore"
author: "Malika Dhakhwa"
date: "2024-02-23"
output: html_document
---

@@

#| message: false
#| warning: false
#load needed packages. 
library(here) #for data loading/saving
library(dplyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(ggplot2) # for visualization
library(tibble) # For rownames_to_column
library(gt) # for table
library(gridExtra)
library(webshot2) #Capture as an Image the gttable before converting it to pdf
library(zoo) #To plot seasonality of sales
```

@@

```{r}
#Path to data using here() package
data_location <- here::here("data","processed-data","processed_superstore_RFM.rds")
#load data
mydata <- readRDS(data_location)
```

@@

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results","tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

# Descriptive Statistics tables

A meaningful analysis of customer spending behavior require the Invoice Sales Amount, total purchase by a Customer during a period, frequency of purchase by a customer during a period or a combination of those. A table is created which provides the descriptive statistics of sales by Invoice value, Customer purchase and frequency of customer purchase. This table reveals that there were 1402 transactions to 399 different customers during 2010-12-01 and	2011-12-09.

@@

# Aggregate total sales by InvoiceNo for calculating mean, standard deviation, minimum and maximum
Sales_by_Invoice <- mydata %>%
    summarise(
    Particulars = "Invoice Value",
    Count = n(),
        Mean = round(mean(InvoiceValue, na.rm = TRUE), 2),
    SD = round(sd(InvoiceValue, na.rm = TRUE), 2),
    Min = round(min(InvoiceValue, na.rm = TRUE), 2),
    Max = round(max(InvoiceValue, na.rm = TRUE), 2)
  )

# Aggregate total sales by CustomerID to calculate mean, standard deviation, minimum and maximum
TotalSales_customer <- mydata %>%
  group_by(CustomerID) %>%
  summarise(TotalSales = sum(InvoiceValue, na.rm = TRUE)) %>%
  ungroup() %>%
  summarise(
    Particulars = "Cumulative Sales to Individual Customers",
    Count = n(),
    Mean = round(mean(TotalSales, na.rm = TRUE), 2),
    SD = round(sd(TotalSales, na.rm = TRUE), 2),
    Min = round(min(TotalSales, na.rm = TRUE), 2),
    Max = round(max(TotalSales, na.rm = TRUE), 2)
  )

# Frequency of sales per CustomerID (i.e., the count of unique InvoiceNo per CustomerID)for calculating mean, standard deviation, minimum and maximum

sales_frequency_per_customer <- mydata %>%
  group_by(CustomerID) %>%
  summarise(FrequencyOfSales = n_distinct(InvoiceNo)) %>%
  ungroup() %>%
  summarise(
    Count = n(),
    Mean = round(mean(FrequencyOfSales, na.rm = TRUE), 2),
    SD = round(sd(FrequencyOfSales, na.rm = TRUE), 2),
    Min = round(min(FrequencyOfSales, na.rm = TRUE), 2),
    Max = round(max(FrequencyOfSales, na.rm = TRUE), 2)
  ) %>%
  mutate(Particulars = "Sales frequency to Individual Customers")

# Combine the statistics into a single data frame
combined_stats <- bind_rows(Sales_by_Invoice, TotalSales_customer, sales_frequency_per_customer)

@@

#Saving this statistics in the results folder.

@@

Next, density plots were created for Sales across customers and Invoice Values. Both figures are heavily right-skewed.The averages of total sales to a customer and total invoice values appear to be less than 2500 and 1250 sterling pounds respectively supporting the mean values of 2025 and 563 in the descriptive table.

A histogram of purchase frequency of each customer is created by grouping each customer by invoice no.The average frequency appears to be less than 5 times for most of the customers which is also observed in the descriptive statistics table.

```{r}
# Aggregate sales by CustomerID
TotalSales_customer <- mydata %>%
  group_by(CustomerID) %>%
  summarise(TotalSales = sum(InvoiceValue, na.rm = TRUE)) %>%
  arrange(desc(TotalSales))

#Density Plot of Sales per Customer
# Assuming sales_per_customer is already calculated
p1 <- ggplot(TotalSales_customer, aes(x = TotalSales)) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  labs(title = "Density Plot of Sales across Customers", x = "Total Sales", y = "Density")

# Histogram of Sales by Invoice values

p2 <- ggplot(mydata, aes(x = InvoiceValue)) +
  geom_density(fill = "lightgreen", alpha = 0.5) +
  labs(title = "Density Plot of  Invoice Sales",
       x = "Invoice Value", 
       y = "Density") +
  theme_minimal()

# Calculate purchase frequency per customer
purchase_frequency <- mydata %>%
  distinct(CustomerID, InvoiceNo) %>% # Ensure unique InvoiceNo per CustomerID
  count(CustomerID, name = "PurchaseFrequency") # Count the number of invoices per customer
library(ggplot2)

# Histogram of purchase frequencies
p3 <- ggplot(purchase_frequency, aes(x = PurchaseFrequency)) +
  geom_histogram(binwidth = 1, fill = "salmon", color = "darkblue") + # Adjust binwidth as needed
  labs(title = "Histogram of Customer Purchase Frequencies",
       x = "No. of Purchases", 
       y = "Count of Customers") +
  theme_minimal()

# Modify p1 to adjust title font size
p1 <- p1 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p1
# Modify p2 to adjust title font size
p2 <- p2 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p2
# Modify p3 to adjust title font size
p3 <- p3 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p3

plot1<-grid.arrange(p1, p2, p3, ncol = 2, nrow=2)

@@

Next, total purchase by a customer was plotted against purchase frequency.A few outliers were noted.

```{r}
#| message: false
#| warning: false
# Calculate purchase frequency and total invoice value per customer
customer_stats <- mydata %>%
  group_by(CustomerID) %>%
  summarise(PurchaseFrequency = n_distinct(InvoiceNo),
            TotalInvoiceValue = sum(InvoiceValue, na.rm = TRUE)) %>%
  ungroup()  # Ensure the data is ungrouped for further analysis

# Scatter plot of Total Invoice Value by Purchase Frequency
p4 <- ggplot(customer_stats, aes(x = PurchaseFrequency, y = TotalInvoiceValue)) +
  geom_point(alpha = 0.5) +  # Adjust alpha for point transparency, if needed
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a linear regression line without standard error
  labs(title = "Total Invoice Value by Purchase Frequency",
       x = "Purchase Frequency",
       y = "Total Invoice Value") +
  theme_minimal()
plot(p4)

figure_file = here("results","figures","Sales_trend.png")
ggsave(filename = figure_file, plot=p4)

Next, Bar chart is created for Sales Distribution by Country. The top five sales are in Germany, France, Spain, Switzerland and Japan.

```{r}
p5 <- mydata %>% 
  group_by(Country) %>% 
  summarise(TotalSales = sum(InvoiceValue)) %>%
  ggplot(aes(x = reorder(Country, TotalSales), y = TotalSales, fill = Country)) + 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  theme(legend.position = "none")

plot(p5)

figure_file = here("results","figures","Countrywise_Sales.png")
ggsave(filename = figure_file, plot=p5)
```
As the store sells gift goods, an inspection is required to check the seasonality of sales. First, monthly total sales is calculated aggregating Invoice Value by months of the Invoice Date.The aggregated sales is then plotted to visualize seasonal trend which depicted a spike in the sales during the last few months of the year.

```{r}
#Aggregating Sales by months
mydata_aggregated <- mydata %>%
  group_by(InvoiceDate = as.yearmon(InvoiceDate)) %>%
  summarize(TotalValue = sum(InvoiceValue))

# Formatting Date for plotting purposes
mydata_aggregated$InvoiceDate <- as.Date(as.yearmon(mydata_aggregated$InvoiceDate))

p6<-ggplot(mydata_aggregated, aes(x = InvoiceDate, y = TotalValue)) +
  geom_line() + 
  scale_x_date(date_labels = "%b", date_breaks = "1 month") + # Adjust formatting as needed
  labs(title = "Monthly Sales", x = "Month", y = "Sales Value") +
  theme_minimal()
p6

@@

>>>>

---
title: "Exploratory Analysis of Superstore"
author: "Malika Dhakhwa"
date: "2024-02-23"
output: html_document
---
The purpose of this task is to perform exploratory analysis of the superstore data.

@@

#| message: false
#| warning: false
#load needed packages. 
library(here) #for data loading/saving
library(dplyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(ggplot2) # for visualization
library(tibble) # For rownames_to_column
library(gt) # for table
library(gridExtra)
library(zoo) #To plot seasonality of sales
library(scales)
```

@@

```{r}
#Path to data using here() package

data_location <- here::here("data","processed-data","processed_superstore.rds")
#load data
superstore <- readRDS(data_location)
```

@@

```{r}
summary_df = skimr::skim(superstore)
print(summary_df)
# save to file
summarytable_file = here("results","tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```
The data summary reveals that there were 14951 transactions to 4117 different customers during 2010-12-01 and	2011-12-09.

# Descriptive Statistics tables
The processed data accounts sales of a particular item as an observation. This means different items sold in the same Invoice appears as separate observations and it is most likely that each Invoice number appeared in separate observations. For descriptive analysis, we need to combine the sales under one invoice as a single observation.

```{r}
#Creating data frame where Sales under a unique Invoice No. is bundled as a single observation
store_InvNo <- superstore %>%
  #grouping the data by the unique InvoiceNo
  group_by(InvoiceNo) %>%
  summarise(InvoiceValue = sum(Sales, na.rm = TRUE), #sums all the sales under the unique InvoiceNo
            InvoiceDate = first(InvoiceDate),  #Selects the first Invoice Date for the data grouped by InvoiceNo 
            CustomerID = first(CustomerID), #Selects the first CustomerID for the data grouped by InvoiceNo
            Country = first(Country)) %>%   ##Selects the first Country for the data grouped by InvoiceNo
select(InvoiceDate, InvoiceNo, CustomerID, InvoiceValue, Country)
str(store_InvNo)
skimr::skim(store_InvNo)
```

For further exploratory analysis, we require customer spending behavior indicators, Recency, Frequency and Monetary. Recency is the number of days since the last purchase by a customer. Frequency is the no. of purchases by a customer over a given period. Monetary is the total amount of purchase a customer has made over a period.

For determining recency of a transaction, a reference date past the study period is required. The data covers transactions till 2011-12-09. The next day i.e. 2011-12-10 is considered as the reference date to measure recency.

```{r}
#fixing the reference date for recency
ref_date <- as.Date("2011-12-10")

#Creating a data frame that includes Recency, frequency and monetary of customer purchase.
store_CustomerID <- superstore %>%
  group_by(CustomerID) %>%
  summarise(Frequency = n_distinct(InvoiceNo),
            Recency = as.numeric(ref_date - as.Date(max(InvoiceDate))),
            Monetary=sum(Sales))%>%
  ungroup()

```
A table is created which provides the descriptive statistics of sales by Invoice value, Customer purchase, frequency and recency of customer purchase.

@@

# Aggregate total sales by InvoiceNo for calculating mean, standard deviation, minimum and maximum
Sales_by_Invoice <- store_InvNo %>%
    summarise(
    Count = n(),
    Particulars = "Invoice Value",
    Mean = round(mean(InvoiceValue, na.rm = TRUE), 2),
    SD = round(sd(InvoiceValue, na.rm = TRUE), 2),
    Min = round(min(InvoiceValue, na.rm = TRUE), 2),
    Max = round(max(InvoiceValue, na.rm = TRUE), 2)
  )

# Aggregate total sales by CustomerID to calculate mean, standard deviation, minimum and maximum
TotalSales_customer <-store_CustomerID%>%
    summarise(
    Count = n(),
    Particulars = "Cumulative Purchase",
    Mean = round(mean(Monetary, na.rm = TRUE), 2),
    SD = round(sd(Monetary, na.rm = TRUE), 2),
    Min = round(min(Monetary, na.rm = TRUE), 2),
    Max = round(max(Monetary, na.rm = TRUE), 2)
  )

# Frequency of sales per CustomerID (i.e., the count of unique InvoiceNo per CustomerID)for calculating mean, standard deviation, minimum and maximum
sales_frequency_per_customer <-store_CustomerID%>%
    summarise(
    Count = n(),
    Mean = round(mean(Frequency, na.rm = TRUE), 2),
    SD = round(sd(Frequency, na.rm = TRUE), 2),
    Min = round(min(Frequency, na.rm = TRUE), 2),
    Max = round(max(Frequency, na.rm = TRUE), 2)
  ) %>%
  mutate(Particulars = "Purchase frequency")

#How Recent a customer bought
Sales_Recency<- store_CustomerID%>%
  summarise(
    Count = n(),
    Particulars = "Recency of Purchase",
    Mean = round(mean(Recency, na.rm = TRUE), 2),
    SD = round(sd(Recency, na.rm = TRUE), 2),
    Min = round(min(Recency, na.rm = TRUE), 2),
    Max = round(max(Recency, na.rm = TRUE), 2)
  )
# Combine the statistics into a single data frame
combined_stats <- bind_rows(Sales_by_Invoice, TotalSales_customer, sales_frequency_per_customer, Sales_Recency)

@@

@@

Next, histograms were created for Sales across customers and Invoice Values. Histograms were also created for purchase frequency and recency of customers.

```{r}
#Histogram of Total Sales per Customer
p1 <- ggplot(store_CustomerID, aes(x = Monetary)) +
  geom_histogram(binwidth = 100,fill = "green", color = "skyblue") +
  labs(title = "Histogram of Customer's Total Purchases", x = "Total Sales", y = "Count")

# Histogram of Sales by Invoice values
p2 <- ggplot(store_InvNo, aes(x = InvoiceValue)) +
  geom_histogram(binwidth = 20, fill = "lightgreen", color = "darkblue") +
  labs(title = "Histogram of  Invoice Value", x = "Invoice Value", y = "Count") +
  theme_minimal()

# Histogram of purchase frequencies
p3 <- ggplot(store_CustomerID, aes(x = Frequency)) +
  geom_histogram(binwidth = 1, fill = "salmon", color = "darkblue") + # Adjust binwidth as needed
  labs(title = "Histogram of Customer Purchase Frequencies",
       x = "No. of Purchases", 
       y = "Count of Customers") +
  theme_minimal()
# Histogram of purchase recencies
p4 <- ggplot(store_CustomerID, aes(x = Recency)) +
  geom_histogram(binwidth = 10, fill = "coral", color = "darkblue") + # Adjust binwidth as needed
  labs(title = "Histogram of Customer Purchase Recency",
       x = "Days since last Purchase", 
       y = "Count of Customers") +
  theme_minimal()

# Modify p1 to adjust title font size
p1 <- p1 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p1
# Modify p2 to adjust title font size
p2 <- p2 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p2
# Modify p3 to adjust title font size
p3 <- p3 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p3
# Modify p4 to adjust title font size
p4 <- p4 + theme(plot.title = element_text(size = 11)) # Adjusting font size of the title of p3

plot1<-grid.arrange(p1, p2, p3,p4, ncol = 2, nrow=2)

@@

```
All histograms exhibited heavily right-skewness. The histogram of Customer's Total Purchases indicated majority of the customers are associated with lower sales amount, typical of individual consumer transactions. Histogram of Invoice value aligns with this finding by showing that most invoices had values consistent with retail purchases. Histogram of Customer Purchase Frequencies indicated that most customers engage in a number of transactions typical for retail shopping behavior. Histogram of Customer Purchase Recency exhibited a higher frequency of recent purchases, which suggest regular shopping habits expected in retail context. ed.

```{r}
#| message: false
#| warning: false

# Scatter plot of Total Invoice Value by Purchase Frequency
p5 <- ggplot(store_CustomerID, aes(x = Frequency, y = Monetary)) +
  geom_point(alpha = 0.5) +  
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add a linear regression line without standard error
  labs(title = "Total Invoice Value by Purchase Frequency",
       x = "Purchase Frequency",
       y = "Cummulative Invoice Value") +
  theme_minimal()
p5

figure_file = here("results","figures","MonetarybyFreq_distribution.png")
ggsave(filename = figure_file, plot=p5)

```
The scatter plot visualizing Total Invoice Value by Purchase Frequency indicates a positive correlation between the two components. This suggests that as the frequency of purchases by a customer increases, so does their total amount spent. Such a finding is typical in retail scenarios where regular customers often contribute more to overall sales than infrequent ones. This trend underscores the importance of strategies aimed at increasing customer engagement and purchase frequency as a means to boost revenue.

As the store is in gift selling business, an inspection is required to check the seasonality of sales. First, monthly total sales is calculated aggregating Invoice Value by months of the Invoice Date.The aggregated sales is then plotted to visualize seasonal trend.

```{r}
#Aggregating Sales by months
superstore_aggregated <- store_InvNo %>%
  group_by(InvoiceDate = as.yearmon(InvoiceDate)) %>%
  summarize(TotalValue = sum(InvoiceValue))

# Formatting Date for plotting purposes
superstore_aggregated$InvoiceDate <- as.Date(as.yearmon(superstore_aggregated$InvoiceDate))

p6<-ggplot(superstore_aggregated, aes(x = InvoiceDate, y = TotalValue)) +
  geom_line() + 
  scale_x_date(date_labels = "%b", date_breaks = "1 month") + 
  scale_y_continuous(labels = label_number()) + # formats the y-axis labels as full numbers
  labs(title = "Monthly Sales", x = "Month", y = "Sales Value") +
  theme_minimal()

p6

@@

```
The data spans from 2010-12-01 and	2011-12-09, revealing a good growth in the December of 2011 compared to the same month previous year. The monthly sales exhibits noteworthy seasonal variation. A sharp peak in May suggests a strong spring season and the plateau from June to August indicates stability in consumer purchases. Starting in October, the sales trajectory showed an upward trend, reaching maximum in November/December, which could be due to early holiday shopping influenced by Black Friday and the approaching Christmas season. It is reasonable to anticipate that past customers will engage in purchasing activities during these peak sales periods.

To gain further insights, an examination of the recency by frequency plot is carried out. This will help to understand the relationship between the recency of purchases and the frequency of customer transactions.

```{r}
# Scatter plot of Recency by Frequency
p7 <- ggplot(store_CustomerID, aes(x = Frequency, y = Recency)) +
  geom_point(alpha = 0.5) +  
  labs(title = "Recency by Frequency Plot",
       x = "Purchase Frequency",
       y = "No. of days since Last Purchase") +
  theme_minimal()
p7

figure_file = here("results","figures","Recency_by_Frequency.png")
ggsave(filename = figure_file, plot=p7)
```
While the Recency by Frequency plot revealed an expected pattern where customers with more frequent purchases tended to have more recent transactions, an unexpected trend was also noted. There are some evidences that a segment of customers with a history of multiple purchases did not return during the peak sales period. This divergence from the usual retail pattern suggests the need of further investigation.

To complement this analysis, another chart is created to examine whether there is a correlation between the recency of purchases and the monetary value of those purchases.

```{r}
# Scatter plot of Recency by Frequency
p8 <- ggplot(store_CustomerID, aes(x = Monetary, y = Recency)) +
  geom_point(alpha = 0.5) +  
  labs(title = "Recency by Monetary plot",
       x = "Monetary",
       y = "No. of days since Last Purchase") +
  theme_minimal()
p8

figure_file = here("results","figures","Recency_by_Monetary.png")
ggsave(filename = figure_file, plot=p8)
```
The Recency by Monetary plot mirrored the trend observed in the 'Recency by Frequency' plot. The plot revealed that a subset of customers who have purchased over 10,000 in monetary value, did not make subsequent purchases within the following 100 days.

The store has customer base inside and outside of the UK. To visualize the sales distribution by countries, a bar chart is plotted.

```{r}
p9 <- store_InvNo %>% 
  group_by(Country) %>% 
  summarise(TotalSales = sum(InvoiceValue)) %>%
  ggplot(aes(x = reorder(Country, TotalSales), y = TotalSales, fill = Country)) + 
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = label_number())+
  coord_flip() + 
  theme(legend.position = "none")

plot(p9)

figure_file = here("results","figures","Countrywise_Sales.png")
ggsave(filename = figure_file, plot=p9)
```
As exhibited, majority of the sales is in UK followed by Germany and France.

To better understand the customer base and identify the most loyal shoppers, the project will focus on customers who have shopped with the store in two distinct periods: the first half of the year and the peak sales season. As revealed by the monthly sales trend, sales during first six months of the store were comparatively slower than the later period of the year. The first half of the year also typically doesn't have big festive drives to attract the buyers. Customers who made purchases during this time might be doing so based on the everyday value and appeal of the products, free from the influence of holiday discount or promotions. During the peak sales season, the shopping activity naturally increases due to holidays surging the store's sales. On the other hand, competition among retailers intensifies, with promotions proliferating across the market and challenging the customer retention. Amidst this competition, customers who chose to return to the same store for their holiday purchases potentially are the ones who shop with the store regularly, regardless of festive season or deal offers. By looking at customers who have made purchases in both of these times, a group that is valuable round the year can be identified. Such group potentially keep steady sales in the slower months and boost the numbers during the busy times. Understanding this group can help focus the marketing efforts, improve the customer service, and ultimately, keep the business strong.

The analysis part of the project is designed to quantify the contributions made by this segment of customers to the store's overall business. Such data-driven understanding is anticipated to help marketing initiatives and customer service enhancements in order to retain loyal customer base.The analysis can be found at R>analysis-code-statistical_analysis.qmd
